`localhierarchy`: an R package to facilitate fitting of global and local Bayesian hierarchical models

### 2025-09-12 paper/paper.qmd
Updated paper with typographical edits; standardized package name on 'localhierarchy'

---
title: "`localHierarchy`: an R package to facilitate fitting of global and local Bayesian hierarchical models"
author: "Leontine, Herb, Shauna, Evan"
format: 
  pdf:
    latex-engine: pdflatex    # lualatex or xelatex or pdflatex
    includes:
      in-header: preamble.tex 
editor: visual
execute:
  cache: true

abstract: | 
  Bayesian hierarchical models are widely used in global health estimation, where data availability may vary across national or subnational populations. Such models are typically fitted to a global data base, e.g., to produce national-level estimates for all countries in the world or in some region. To facilitate analysis at a local level, models are often desirable that are informed by global models but can be fitted to just a subset of the data, e.g., data for one country. We refer to such models as local models; models that are derived from global hierarchical models but adapted such that they can be fitted to the data from one population alone.

  We present the `localHierarchy` R package, which provides functionality for fitting Bayesian hierarchical models in settings where both global and local estimation is required. The package provides R functions and Stan model components to support global modeling, in which all parameters in hierarchical models are estimated, and local modeling, where parameters are estimated for only one or a small number of populations, using fixed values from a global model fit.

@@

# Introduction
see overleaf

@@

The modeling approach introduced in `localhierarchy` was motivated by the hierarchical structures associated with national and subnational level estimation. These structures are illustrated in Figure 1, showing how countries are grouped within regions, regions are grouped in major world regions.

In this section, we introduce the main set up used for global and local hierarchical models in this context of national estimation. We use general notation and a specific example focused on national level estimation. Applications and an extension to subnational estimation are included in the use cases.

@@

Suppose we want to estimate some population parameter of interest $\mu_r$ for a population indexed by $r$. For example, $\mu$ could refer to some population health indicator such as service coverage or morbidity or mortality-related outcomes. Index $r$ refers to the population in a region indexed by $r$, for example, a subnational region, a country, or a group of countries. We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model set up, see Figure 2: \begin{align*}
\mu_{r^{(l)}}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma^{(l)}_\theta &\sim N(\mu^{(l-1)}_{r^{(l-1)}}, \sigma^{(l)}_\mu), \text{ for } l > 0,
\end{align*} where $\mu_{r^{(l)}}^{(l)}$ ... Here, $\mu^{(l-1)}_{r^\prime}$ is the parameter value at one level higher in the hierarchy for the region $r^\prime$ that contains unit $r$. The parameter $\sigma_\mu^{(l)}$ for $l > 0$ describes variability of values of the parameter $\mu$ across geographic units at level $l$ of the hierarchy. At the top level of the hierarchy, $l = 0$, only a single parameter is estimated.

@@

As a specific example, we consider the estimation of country-specific parameters, where countries are organized into subregions of countries as illustrated in Figure 1. In this setting, using alternative notation for the same set up to indicate the level by name, the hierarchical model can be written as follows:

\begin{align}
\mu_c|\mu_{s[c]}^{(subregion)}, \sigma_{\mu, country} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{\mu, country}^2), \label{eq-countryhier} \\
\mu_s^{(subregion)}|\mu_{w[s]}^{(region)}, \sigma_{\mu, subregion}^2 &\sim N(\mu_{r[s]}^{(region)}, \sigma_{\mu, subregion}^2),\nonumber \\
\mu_w^{(region)}|\mu^{(global)}, \sigma_{\mu, regions}^2 &\sim N(\mu^{(global)}, \sigma_{\mu, region}^2), \nonumber
\end{align} where $c =1, ..., C$ refers to the country index, $s[c] = 1, .., S$ to the subregion index of country $c$, and $w[s[c]] = 1, .., W$ to the world region index of subregion $s[c]$. The parameter $\mu^{(global)}$ is the global mean for the parameter of interest, e.g., a global mean for service coverage or mortality. The parameters $\sigma_{\mu, country}$, $\sigma_{\mu, subregions}$, and $\sigma_{\mu, regions}$ describe variability in the parameter of interest across countries, subregions, and regions, respectively.

<!-- Here, if $\mu^{(global)}$ has prior $\mu^{(global)} \sim N(0, s_{\mu, global}^2)$, then $$\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(region)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c},$$ with $\gamma \sim N(0,1)$ for each gamma (one global, $R$ regions, $S$ subregions, $C$ countries). -->

@@

A local model is derived from a global one by fixing parameters. Continuing with the specific example, an example local model to produce country estimates using data from one country alone has the following set up:

$$\mu_c \sim N(\hat{\mu}_{s[c]}^{(subregion)}, \hat{\sigma}_{\mu, country}^2),$$

@@

The `localhierarchy` R package contains R functions and example Stan model files to fit global and local hierarchical models. The package is set up to allow for users to incorporate the global-local hierarchy set up into their own project. For a given parameter and hierarchical model specification, the package functions enable processing of data to produce Stan input data for global and local models, Stan model components that work for both global as well as local models, and R functions for postprocessing of model outputs to check results and produce summary files of global models for use in local models.

In a typical work flow, a user first fits a global model, followed by local ones. Worked examples are given in the use case section. The global and local models in these examples are fitted using the `localhierarchy` wrapper function `fit_model_localhierarchy`, using its arguments to specify a hierarchical model, data, and, for local fits, pass on outputs from the global model. The wrapper function calls R subfunctions to process data and prepare Stan input data, read in a Stan model, fit the model, and to postprocess the results.

The package includes a user-friendly specification of hierarchical model structures, and associated specification of fixed parameters for local runs. A worked example is given in the first use case. Briefly, if a user wants to specify a hierarchical model like the one introduced above, with levels world, regions, subregion, and country, <!-- country parameters $\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(region)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c}$,  --> using a data set with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows \[Q for herb: refs/credit to be given for this set up to other work?\]:

@@

The parametrization used in Stan is a non-centered parametrization. In general notation, for $l>0$, non-centered parametrization refers to writing mean parameters as deviations from higher-level group means: \begin{align*}
\mu_r^{(l)} &= \mu^{(l-1)} + \sigma^{(l)} \cdot \gamma_{r^\prime}^{(l-1)},
\end{align*} where $\gamma_r^{(l)} \sim N(0, 1)$. Applied sequentially, and if $\mu^{(0)} \sim N(0, s)$, we get \begin{align*}
\mu_r^{(l)} &= s\cdot\gamma^{(0)} + \sum_{k=1}^{l} \sigma^{(k)} \cdot \gamma_{r^\prime}^{(k)},
\end{align*} where $r^\prime$ now refers more generally to the index of the unit containing the lower-level group. For the specific example of country parameters, the non-centered parametrization is given by $$\mu_c =  s_{global}\cdot {\gamma}_{global} + {\sigma}_{\mu, region}\cdot {\gamma}_{r[s[c]]]}^{(region)} + {\sigma}_{\mu, subregion}\cdot{\gamma}_{s[c]}^{(subregion)} + {\sigma}_{\mu, country}\cdot\gamma_{c}.$$

Implementation is based on a model matrix. For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c$s: $$(\mu_1, ..., \mu_c) = M\cdot (\eta_1, ..., \eta_{1+R+S+C}),$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\gamma_1 = (s_{global}\cdot \gamma_{global}, \gamma_2 = \sigma_{\mu, region}\cdot\gamma_{\mu, 1}^{(region)},..., \gamma_{1+W+S+C} = \sigma_{\mu, country}\cdot\gamma_{\mu, C}$. When estimating a subset of parameters in a local model, fixed values of $\sigma$s and $\gamma$s by point estimates (posterior means) from a global fit. For example, for a national model, we get $$\mu_c =  s_{global}\cdot \hat{\gamma}_{global} + \hat{\sigma}_{\mu, region}\cdot \hat{\gamma}_{r[s[c]]]}^{(region)} + \hat{\sigma}_{\mu, subregion}\cdot\hat{\gamma}_{s[c]}^{(subregion)} + \hat{\sigma}_{\mu, country}\cdot\gamma_{c}.$$

@@

## Simulation set up

The simulation set up is summarized in Figure 4. It covers both national and subnational parameters and data. While for national level estimation alone, the subnational settings are not needed, we add them here to create a set up that can serve to illustrate both national as well as subnational-level estimation.

@@

The top box in Figure 4 indicates the hierarchical model considered. For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c|\mu^{(global)}, \sigma_{\mu, country} \sim N(\mu^{(global)}, \sigma_{\mu, country}^2)$. For subnational estimation, an additional level is added, with $\mu_r^{(subnat)} =  \mu_{c[r]} + \eta_{r}^{(subnat)}$, $\eta_{r}^{(subnat)}|\sigma_{\mu, subnat} \sim N(0, \sigma_{\mu, subnat})$.

We chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality, as opposed to the data model aspects of modeling. In practice, the data model can be set by the user to reflect actual data generating mechanisms, for example, by using a binomial likelihood, considering transformed outcomes, or accounting for sampling errors for observations that were obtained from a complex survey design.

For generating national level data (box 2), we assume a normal likelihood with constant variance: $$y_i^{(country)}|\mu_{c[i]}, \sigma_y^2 \sim N(\mu_{c[i]}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$. Again, the use of a normal likelihood with constant variance is for illustration purposes only. In practice, the data model can be set by the user to reflect the observational error. We also generate subnational data following a normal distribution: $$y^{(subnat)}_i|\mu^{(subnat)}_{g[i]}, \sigma_y^2 \sim N(\mu^{(subnat)}_{r[i]}, \sigma_y^2),$$ where $g[i]$ refers to the subnational region of subnational observation $i$. In this simulation set up, we generate subnational data independently of national-level data, and we use the same constant data variance at the national and subnational levels. In real examples, data variances are likely to differ (for example, because subnational sampling errors are larger at the subnational level) and national and subnational data may be derived from the same source. We discuss an example of those settings in the case study.

We create simulated data sets using the `simulate_multilevel_data` function. We simulate a set of model parameters, setting $\mu^{(global)} = 0$, $\sigma_{\mu,country} = 0.5$, $\sigma_{\mu}^{(subnat)} = 0.2$, creating 30 countries with 20 subnational regions in each country. We then simulate subnational data, using $\sigma_y = 0.1$ and generating 10 observations per region. Finally, the function adds national data as well using the argument `add_data1levelup`, to output the subnational and national level data sets:

```{r}
dat_all <- simulate_multilevel_data(
  n_levels = 2,
  n_units_perlevel = c(30, 20),
  sigma_perlevel = c(0.5, 0.2),
  mu_global = 0,
  add_data = TRUE,
  n_data = 10,
  sigma_y = 0.1,
  add_data1levelup = TRUE)
dat_subnat <- dat_all$data
dat_nat <- dat_all$data_1levelup
```

In the simulated data, level1 refers to the country level and level2 refers to the subnational level. The data sets contain the simulated parameters (`eta`) and data (`y`):

```{r}
dat_subnat %>% select(level1, level2, mu_global, level1_eta, level2_eta, y)
dat_nat %>% select(level1,  mu_global, level1_eta, y)
```

@@

```{r}
geo_unit_index <- get_geo_unit_index_data(dat_nat,
                                  hierarchical_levels = hierarchical_level,
                                  area = "level1")
geo_unit_index
```

The second step in the processing is to create a hierarchical data object, using the function `hierarchical_data`. This function outputs a list with various components associated with the parametrization of the hierarchical parameters, used in the Stan model and for further post-processing:

@@

All this information is passed to the function `hierarchical_param_stan_data` to produce inputs needed in the Stan model fitting. This function outputs a named list with Stan input data relevant to the hierarchical set up:

@@

For example, in this setting, we only have to estimate one $\sigma_{\mu}$, and no $\sigma_{\mu}$ is fixed, as recorded in the `n_sigma_fixed` and `n_sigma_estimate` elements of the Stan input data:

```{r}
hier_stan_data$mu_n_sigma_fixed
hier_stan_data$mu_n_sigma_estimate
```

Excerpts from the Stan model are copied below to introduce model specification. The same Stan model is used for both global and local models. The first excerpt starts at the end in terms of model specification and concerns the specification of the hierarchical parameters $\mu_c$, using the `localhierarchy` Stan function `get_mu`:

@@

Here we use that $(\mu_1, ..., \mu_c) = M\cdot (\eta_1, ..., \eta_c)$, where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_c)$ refers to the vector with products of standard deviation and $\gamma$ terms. The function `get_mu` calculates the vector of $\mu_c$s from the vector of $\eta$s (`mu_star` in the Stan model) and a decomposition of $M$ into different parts $U$, $V$, $W$. This implementation uses the sparse representation of the model matrix $M$ with Stan functions `csr_matrix_times_vector` and `crs_extract`.

@@

In the model code, the data block contains the input data for the $\gamma$s and $\sigma$s. The `localhierarchy` Stan function `get_mu_star` is used to obtain the $\eta$s. Its arguments also include prior parameters `mu_scalarprior_mean` and `mu_scalarprior_sd`, which refer to the (rescaled) prior mean and standard deviation of $\mu^{(global)}$. The model block contains the priors for the $\gamma$s and $\sigma_{\mu}$s, given by \begin{align*}
\mu^{(global)} &\sim N(0,1),\\
\sigma_{\mu} &\sim \text{Half-Normal}(0, s_{\mu}).
\end{align*} All prior parameters are added as input data to the Stan model.

@@

```{r}
global_fit <- fit_model_localhierarchy(runstep = "global_national",
                                       hierarchical_level = hierarchical_level,
                                       area = "level1",
                                       survey_df = dat_nat)
```

@@

```{r}
global_fit$post_summ %>% 
  filter(variable_no_index == "mu_sigma")
```

We can do a local fit by passing on the global fit object to the wrapper function and including data for countries of interest. Here we do local runs for all countries, by including the entire data set, and specify what to fix using the `runstep` argument:

```{r}
fit_local <- fit_model_localhierarchy(runstep = "local_national",
                                      global_fit = global_fit,
                                      area = "level1",
                                      survey_df = dat_nat)
```

@@

Above column `quant` refers to posterior means and upper and lower bounds of 95% credible intervals. We can create plots to show outputs, for example for the first 4 countries:

```{r}
plot_posterior_summaries(res = res_global, modelname1 = "global",
                          hierarchy_select = "level1",
                          areas_select = unique(res_local$level1$name)[1:4]
                                      )$level1
```

@@

```{r}
plot_posterior_summaries(res = res_local, modelname1 = "local",
                         res2 = res_global, modelname2 = "global", 
                         hierarchy_select = "level1",
                         areas_select = unique(res_local$level1$name)[1:4]
                                      )$level1
```

@@

```{r}
plot_posterior_summaries(res = res_local, modelname1 = "local",
                         res2 = res_global, modelname2 = "global")$intercept 
```

@@

### Workflow for nationally-informed subnational estimation
For many global health applications, subnational estimation is of interest to inform policy and programmatic decision making at a more local level. If data are limited, models can be considered to share information across regions. We consider here the use of hierarchical models for a parameter of interest. The `localhierarchy` package can be used to fit subnational hierarchical models, following the same principles as for national estimation or through an updated approach. Just as for national-level hierarchical modeling, the starting point for subnational hierarchical modeling is to specify a hierarchical modeling structure for the subnational regions, such as the one from Figure 1. When following the same principles as for national-level hierarchical modeling, a global model can be fitted to a global data base of subnational data, followed by local models for subnational regions.

In addition to a standard approach to subnational modeling, functionality in the `localhierarchy` package can also be used to fit subnational models that are informed by national level data and estimation, referred to here as "nationally-informed subnational estimation (NISE)". This approach is useful when subnational data are limited, for example, in terms of the number of countries with subnational data, the number of subnational regions with data, the number of observations per subnational region, or due to uncertainty associated with subnational data. An example is given in the case study on family planning estimation, where national level data are available for a large number of country-years while a global subnational data set is more limited in terms of the number of country-years included.

@@

```{r}
area_subnat <- "level2"
global_fitsubnat_seq <- fit_model_localhierarchy(
   runstep = "global_subnational",
   area = area_subnat, # subnational area name 
   survey_df = dat_subnat, # pass on subnat data
   # for NISE:
   global_fit = global_fit, # global national fit
   add_subnational_hierarchy = area_subnat
   # what level to add to the hierarchy, compared to the 
   # global national model 
   )
```

@@

```{r}
global_fit$post_summ %>%
  filter(str_detect(variable_no_index, "sigma"))
global_fitsubnat_seq$post_summ %>%
  filter(str_detect(variable_no_index, "sigma"))
```

@@

```{r}
fit_local_seq <- fit_model_localhierarchy(runstep = "local_subnational",
                                   area = area_subnat,
                                   survey_df = dat_subnat,
                                   global_fit = global_fitsubnat_seq)
```

>>>>

---
title: "`localhierarchy`: an R package to facilitate fitting of global and local Bayesian hierarchical models"
author: "Leontine, Herb, Shauna, Evan"
format: 
  pdf:
    latex-engine: pdflatex    # lualatex or xelatex or pdflatex
    include-in-header: 
      - file: preamble.tex 
editor: visual
execute:
  cache: true

abstract: | 
  Bayesian hierarchical models are widely used in global health estimation, where data availability may vary across national or subnational populations. Such models are typically fitted to a global database, e.g., to produce national-level estimates for all countries in the world or in some region. To facilitate analysis at a local level, models are often desirable that are informed by global models but can be fitted to just a subset of the data, such as data for one country. We refer to such models as local models: models that are derived from global hierarchical models but adapted such that they can be fitted to the data from one population alone.

  We present the `localHierarchy` R package, which provides functionality for fitting Bayesian hierarchical models in settings where both global and local estimation is required. The package provides R functions and Stan model components to support global modeling, in which all parameters in hierarchical models are estimated, and local modeling, in which parameters are estimated for only one or a small number of populations, using fixed values from a global model fit.

@@

# Introduction

@@

The modeling approach introduced in `localhierarchy` was motivated by the hierarchical structures associated with national and subnational level estimation. These structures are illustrated in Figure 1, showing how countries are grouped within regions, and regions are grouped in major world regions.

In this section, we introduce the principle setup used for global and local hierarchical models in the context of national estimation. We use general notation and a specific example focused on national level estimation. Applications and an extension to subnational estimation are included in the use cases.

@@

Suppose we want to estimate some population parameter of interest $\mu_r$ for a population indexed by $r$. For example, $\mu$ could refer to some population health indicator such as service coverage or morbidity or mortality-related outcomes. Index $r$ refers to the population in a region indexed by $r$, for example, a subnational region, a country, or a group of countries. We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure 2: \begin{align*}
  \mu_{r^{(l)}}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma^{(l)}_\theta &\sim N(\mu^{(l-1)}_{r^{(l-1)}}, \sigma^{(l)}_\mu), \text{ for } l > 0,
\end{align*} 
where $\mu_{r^{(l)}}^{(l)}$ ... Here, $\mu^{(l-1)}_{r^\prime}$ is the parameter value at one level higher in the hierarchy for the region $r^\prime$ that contains unit $r$. The parameter $\sigma_\mu^{(l)}$ for $l > 0$ describes variability of values of the parameter $\mu$ across geographic units at level $l$ of the hierarchy. At the top level of the hierarchy, $l = 0$, only a single parameter is estimated.

@@

As a specific example, we consider the estimation of country-specific parameters, where countries are organized into subregions of countries (illustrated in Figure 1). In this setting, using alternative notation for the same setup to indicate the level by name, the hierarchical model can be written as follows:

\begin{alignat}{2}
  \mu_c &\mid \mu_{s[c]}^{(\subregion)}, \sigma_{\mu, \country} &&\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{\mu, \country}^2), \label{eq-countryhier} \\
  \mu_s^{(\subregion)} &\mid \mu_{w[s]}^{(\region)}, \sigma_{\mu, \subregion}^2 &&\sim N(\mu_{r[s]}^{(\region)}, \sigma_{\mu, \subregion}^2), \nonumber \\
  \mu_w^{(\region)} &\mid \mu^{(\glob)}, \sigma_{\mu, \region}^2 &&\sim N(\mu^{(\glob)}, \sigma_{\mu, \region}^2), \nonumber
\end{alignat} 
where $c =1,\dots,C$ refers to the country index, $s[c] = 1, .., S$ to the subregion index of country $c$, and $w[s[c]] = 1, .., W$ to the region index of subregion $s[c]$. The parameter $\mu^{(\glob)}$ is the global mean for the parameter of interest, e.g., a global mean for service coverage or mortality. The parameters $\sigma_{\mu, \country}$, $\sigma_{\mu, \subregion}$, and $\sigma_{\mu, \region}$ describe variability in the parameter of interest across countries, subregions, and regions, respectively.

<!-- Here, if $\mu^{(\glob)}$ has prior $\mu^{(\glob)} \sim N(0, s_{\mu, global}^2)$, then $$\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(
egion)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c},$$ with $\gamma \sim N(0,1)$ for each gamma (one global, $R$ regions, $S$ subregions, $C$ countries). -->

@@

A local model is derived from a global one by fixing parameters. Continuing with the specific example, an example local model to produce country estimates using data from one country alone has the following setup:

$$\mu_c \sim N(\hat{\mu}_{s[c]}^{(\subregion)}, \hat{\sigma}_{\mu, \country}^2),$$

@@

The `localhierarchy` R package contains R functions and example Stan model files to fit global and local hierarchical models. The package is set up to allow for users to incorporate the global-local hierarchy setup into their own project. For a given parameter and hierarchical model specification, the package functions enable processing of data to produce Stan input data for global and local models, Stan model components that work for both global as well as local models, and R functions for postprocessing of model outputs to check results and produce summary files of global models for use in local models.

In a typical work flow, a user first fits a global model, followed by local ones. Worked examples are given in the use case section. The global and local models in these examples are fitted using the `localhierarchy` wrapper function `fit_model_localhierarchy`, using its arguments to specify a hierarchical model, data, and, for local fits, pass on outputs from the global model. The wrapper function calls R functions to process data and prepare Stan input data, read in a Stan model, fit the model, and to postprocess the results.

The package includes a user-friendly specification of hierarchical model structures, and associated specification of fixed parameters for local runs. A worked example is given in the first use case. Briefly, if a user wants to specify a hierarchical model like the one introduced above, with the levels world, regions, subregion, and country, <!-- country parameters $\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(
egion)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c}$,  --> using a dataset with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows \[Q for herb: refs/credit to be given for this set up to other work?\]:

@@

The parameterization used in Stan is a non-centered parameterization. In general notation, for $l>0$, non-centered parameterization refers to writing mean parameters as deviations from higher-level group means: 
\begin{align*}
  \mu_r^{(l)} &= \mu^{(l-1)} + \sigma^{(l)} \cdot \gamma_{r^\prime}^{(l-1)},
\end{align*} 
where $\gamma_r^{(l)} \sim N(0, 1)$. Applied sequentially, and if $\mu^{(0)} \sim N(0, s)$, we get 
\begin{align*}
  \mu_r^{(l)} &= s\cdot\gamma^{(0)} + \sum_{k=1}^{l} \sigma^{(k)} \cdot \gamma_{r^\prime}^{(k)},
\end{align*}
where $r^\prime$ now refers more generally to the index of the unit containing the lower-level group. For the specific example of country parameters, the non-centered parameterization is given by 
\begin{align*}
  \mu_c =  s_{\glob} \cdot {\gamma}_{\glob} + {\sigma}_{\mu, \region}\cdot {\gamma}_{r[s[c]]]}^{(\region)} + {\sigma}_{\mu, \subregion}\cdot{\gamma}_{s[c]}^{(\subregion)} + {\sigma}_{\mu, \country}\cdot\gamma_{c}.
\end{align*}

Implementation is based on a model matrix. For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c$s: $$(\mu_1, ..., \mu_c) = M\cdot (\eta_1, ..., \eta_{1+R+S+C}),$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\gamma_1 = (s_{\glob}\cdot \gamma_{\glob}, \gamma_2 = \sigma_{\mu, \region} \cdot \gamma_{\mu, 1}^{(\region)},..., \gamma_{1+W+S+C} = \sigma_{\mu, \country}\cdot\gamma_{\mu, C}$. When estimating a subset of parameters in a local model, fixed values of $\sigma$s and $\gamma$s by point estimates (posterior means) from a global fit. For example, for a national model, we get 
\begin{align*}
  \mu_c = s_{\glob} \cdot \hat{\gamma}_{\glob} + 
          \hat{\sigma}_{\mu, \region} \cdot \hat{\gamma}_{r[s[c]]]}^{(\region)} + 
          \hat{\sigma}_{\mu, \subregion} \cdot \hat{\gamma}_{s[c]}^{(\subregion)} + 
          \hat{\sigma}_{\mu, \country} \cdot \gamma_{c}.
\end{align*}

@@

## Simulation setup

The simulation setup is summarized in Figure 4. It covers both national and subnational parameters and data. While for national level estimation alone, the subnational settings are not needed, we add them here to create a setup that can serve to illustrate both national as well as subnational-level estimation.

@@

The top box in Figure 4 indicates the hierarchical model considered. For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c|\mu^{(\glob)}, \sigma_{\mu, \country} \sim N(\mu^{(\glob)}, \sigma_{\mu, \country}^2)$. For subnational estimation, an additional level is added, with $\mu_r^{(\subnat)} = \mu_{c[r]} + \eta_{r}^{(\subnat)}$, $\eta_{r}^{(\subnat)}|\sigma_{\mu, \subnat} \sim N(0, \sigma_{\mu, \subnat})$.

We chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling. In practice, the data model can be set by the user to reflect actual data generating mechanisms, for example, by using a binomial likelihood, considering transformed outcomes, or accounting for sampling errors for observations that were obtained from a complex survey design.

For generating national level data (box 2), we assume a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}, \sigma_y^2 \sim N(\mu_{c[i]}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$. Again, the use of a normal likelihood with constant variance is for illustration purposes only. In practice, the data model can be set by the user to reflect the observational error. We also generate subnational data following a normal distribution: $$y^{(subnat)}_i|\mu^{(subnat)}_{g[i]}, \sigma_y^2 \sim N(\mu^{(subnat)}_{r[i]}, \sigma_y^2),$$ where $g[i]$ refers to the subnational region of subnational observation $i$. In this simulation setup, we generate subnational data independently of national-level data, and we use the same constant data variance at the national and subnational levels. In real examples, data variances are likely to differ (for example, because subnational sampling errors are larger at the subnational level) and national and subnational data may be derived from the same source. We discuss an example of those settings in the case study.

Simulated data can be generated using  `simulate_multilevel_data` function. We simulate a set of model parameters, setting $\mu^{(\glob)} = 0$, $\sigma_{\mu,\country} = 0.5$, $\sigma_{\mu}^{(\subnat)} = 0.2$, creating 30 countries with 20 subnational regions in each country. We then simulate subnational data, using $\sigma_y = 0.1$ and generating 10 observations per region. Finally, the function adds national data as well using the argument `add_data1levelup`, to output the subnational and national level data sets:

```{r}
dat_all <- simulate_multilevel_data(
  n_levels = 2,
  n_units_perlevel = c(30, 20),
  sigma_perlevel = c(0.5, 0.2),
  mu_global = 0,
  add_data = TRUE,
  n_data = 10,
  sigma_y = 0.1,
  add_data1levelup = TRUE
)

dat_subnat <- dat_all$data
dat_nat <- dat_all$data_1levelup
```

In the simulated data, `level1` refers to the country level and `level2` refers to the subnational level. The data sets contain the simulated parameters (`eta`) and data (`y`):

```{r}
dat_subnat |> select(level1, level2, mu_global, level1_eta, level2_eta, y)
dat_nat |> select(level1,  mu_global, level1_eta, y)
```

@@

```{r}
geo_unit_index <- get_geo_unit_index_data(
  dat_nat,
  hierarchical_levels = hierarchical_level,
  area = "level1"
)
geo_unit_index
```

The second step in the processing is to create a hierarchical data object, using the function `hierarchical_data`. This function outputs a list with various components associated with the parameterization of the hierarchical parameters, used in the Stan model and for further post-processing:

@@

All this information is passed to the function `hierarchical_param_stan_data` to produce inputs needed in the Stan model fitting. This function outputs a named list with Stan input data relevant to the hierarchical setup:

@@

For example, in this setting, we only have to estimate one $\sigma_{\mu}$, and no $\sigma_{\mu}$ is fixed, as recorded in the `n_sigma_estimate` and `n_sigma_fixed` elements of the Stan input data:

```{r}
hier_stan_data$mu_n_sigma_estimate
hier_stan_data$mu_n_sigma_fixed
```

Excerpts from the Stan model are copied below to introduce the model specification. The same Stan model is used for both global and local models. The first excerpt starts at the end in terms of model specification and concerns the specification of the hierarchical parameters $\mu_c$, using the `localhierarchy` Stan function `get_mu`:

@@

Here we use that $(\mu_1, ..., \mu_c) = M \cdot (\eta_1, ..., \eta_c)$, where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_c)$ refers to the vector with products of standard deviation and $\gamma$ terms. The function `get_mu` calculates the vector of $\mu_c$s from the vector of $\eta$s (`mu_star` in the Stan model) and a decomposition of $M$ into different parts $U$, $V$, $W$. This implementation uses the sparse representation of the model matrix $M$ with Stan functions `csr_matrix_times_vector` and `crs_extract`.

@@

In the model code, the data block contains the input data for the $\gamma$s and $\sigma$s. The `localhierarchy` Stan function `get_mu_star` is used to obtain the $\eta$s. Its arguments also include prior parameters `mu_scalarprior_mean` and `mu_scalarprior_sd`, which refer to the (rescaled) prior mean and standard deviation of $\mu^{(\glob)}$. The model block contains the priors for the $\gamma$s and $\sigma_{\mu}$s, given by 
\begin{align*}
  \mu^{(\glob)} &\sim N(0,1),\\
  \sigma_{\mu} &\sim \text{Half-Normal}(0, s_{\mu}).
\end{align*} 
All prior parameters are added as input data to the Stan model.

@@

```{r}
global_fit <- fit_model_localhierarchy(
  runstep = "global_national",
  hierarchical_level = hierarchical_level,
  area = "level1",
  survey_df = dat_nat
)
```

@@

```{r}
global_fit$post_summ |> 
  filter(variable_no_index == "mu_sigma")
```

A local fit can be esetimated by passing on the global fit object to the wrapper function and including data for countries of interest. Here we do local runs for all countries, by including the entire data set, and specify what to fix using the `runstep` argument:

```{r}
fit_local <- fit_model_localhierarchy(
  runstep = "local_national",
  global_fit = global_fit,
  area = "level1",
  survey_df = dat_nat
)
```

@@

The column `quant` above refers to posterior means and upper and lower bounds of 95% credible intervals. We can create plots to show outputs, for example for the first 4 countries:

```{r}
plots <- plot_posterior_summaries(
  res = res_global, 
  modelname1 = "global",
  hierarchy_select = "level1",
  areas_select = unique(res_local$level1$name)[1:4]
)

plots$level1
```

@@

```{r}
plots <- plot_posterior_summaries(
  res = res_local, 
  modelname1 = "local",
  res2 = res_global, 
  modelname2 = "global", 
  hierarchy_select = "level1",
  areas_select = unique(res_local$level1$name)[1:4]
)

plots$level1
```

@@

```{r}
plots <- plot_posterior_summaries(
  res = res_local, 
  modelname1 = "local",
  res2 = res_global, 
  modelname2 = "global"
)

plots$intercept 
```

@@

### Workflow for nationally-informed subnational estimation

For many global health applications, subnational estimation is of interest to inform policy and programmatic decision making at a more local level. If data are limited, models can be considered to share information across regions. We consider here the use of hierarchical models for an indicator of interest. The `localhierarchy` package can be used to fit subnational hierarchical models, following the same principles as for national estimation or through an updated approach. Just as for national-level hierarchical modeling, the starting point for subnational hierarchical modeling is to specify a hierarchical modeling structure for the subnational regions, such as the one from Figure 1. When following the same principles as for national-level hierarchical modeling, a global model can be fitted to a global data base of subnational data, followed by local models for subnational regions.

In addition to a standard approach to subnational modeling, functionality in the `localhierarchy` package can also be used to fit subnational models that are informed by national level data and estimation, referred to here as _nationally-informed subnational estimation_ (NISE). This approach is useful when subnational data are limited, for example, in terms of the number of countries with subnational data, the number of subnational regions with data, the number of observations per subnational region, or due to uncertainty associated with subnational data. An example is given in the case study on family planning estimation, where national level data are available for a large number of country-years while a global subnational data set is more limited in terms of the number of country-years included.

@@

```{r}
area_subnat <- "level2"
global_fitsubnat_seq <- fit_model_localhierarchy(
   runstep = "global_subnational",
   area = area_subnat, # subnational area name 
   survey_df = dat_subnat, # pass on subnat data
   # for NISE:
   global_fit = global_fit, # global national fit
   # what level to add to the hierarchy, compared to the 
   # global national model 
   add_subnational_hierarchy = area_subnat
  )
```

@@

```{r}
global_fit$post_summ |>
  filter(str_detect(variable_no_index, "sigma"))
global_fitsubnat_seq$post_summ |>
  filter(str_detect(variable_no_index, "sigma"))
```

@@

```{r}
fit_local_seq <- fit_model_localhierarchy(
  runstep = "local_subnational",
  area = area_subnat,
  survey_df = dat_subnat,
  global_fit = global_fitsubnat_seq
)
```
